  
---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

1. Reading the data
```{r}
data <- read.csv("activity.csv")
```

2. Transform the variable date with date format
```{r}
data$date <- as.Date(data$date, "%Y-%m-%d")
```


## Total number of steps taken per day
```{r}
# Filter NAs
library(dplyr)
good <- !(is.na(data$steps))
data2 <- data[good,]
```

1. Total number of steps taken per day
```{r}
steps_total <- data2 %>% group_by(date) %>% summarise(total = sum(steps))
steps_total
```


2. Histogram of the total number of steps taken per day
```{r}
hist(steps_total$total, xlab = "Steps per day", main = "Histogram of total number steps taken per day", col = "dark green", breaks = 20)
```


3. Obtaining the mean and the median number of steps by day
```{r}
agg <- (data2 %>% group_by(date) 
    %>% summarise(mean_steps = mean(steps), median_steps = median(steps)) 
    %>% print)
```


## Daily activity pattern


1. Creating a series plot


First, the data needs to be aggregated by the mean for each interval
```{r}
library(ggplot2)
data_plot <- (data2 %>% group_by(interval) 
    %>% summarise(mean_steps = mean(steps)) 
    %>% print)

# Creating the plot
g <- ggplot(data_plot, aes(x = interval, y = mean_steps))
g + geom_line(color = "blue", size = 1) + labs(title = "Average daily activity pattern by intervals", x = "Interval", y = "Average steps")
```


2. Getting the interval with max number of steps
```{r}
n <- (data2 %>% group_by(interval) 
    %>% summarise(mean_steps = mean(steps))
    %>% summarise(max_n = which.max(mean_steps))
    %>% print)
max_num <- data2$interval[n$max_n]
```


On average across all the days in the dataset, the 5 minute interval that contains the maximum number of steps is `r max_num`


## Imputing missing values


1. Getting total missing values
```{r}
missing <- sum(is.na(data))
missing
```
The total number of missing values in the dataset is `r missing`


2. Using the steps mean by interval to fill the observations with missing values
```{r}
length(which(data$interval == 0)) # There are 61 repeated intervals
```


3. Creating a new dataset with the missing data filled in
```{r}
# New dataset
new_data <- data
# Using the previous data to obtain the mean steps by interval
new_data$mean_steps <- rep(data_plot$mean_steps, 61)

# loop to fill missing values
for (i in 1:nrow(new_data)) {
    if (is.na(new_data$steps[i])){
        new_data$steps[i] <- new_data$mean_steps[i]
    } else{next}
}
```


4. Obtaining a histogram with the new dataset
```{r}
totals <- new_data %>% group_by(date) %>% summarise(total = sum(steps))
with(totals, hist(total, xlab = "Steps per day", main = "Total number steps taken per day", col = "blue", breaks = 20))

agg2 <- (new_data %>% group_by(date) 
    %>% summarise(mean_steps = mean(steps), median_steps = median(steps)) 
    %>% print)
```


## Are there differences in activity patterns between weekdays and weekends?


1. Creating a new factor variable with two levels: "weekday" and "weekend"
```{r}
library(lubridate)
week1 = weekdays(data$date)

# Transforming the variable weekdays into two levels
week1[week1 %in% c("lunes","martes","miércoles","jueves","viernes")] <- "weekday"
week1[week1 %in% c("sábado","domingo")] <- "weekend"

new_data$weekdays <- factor(week1)
str(new_data)
```


2. Creating a series plot by each level
```{r}
data_plot2 <- (new_data %>% group_by(weekdays,interval) 
    %>% summarise(mean_steps = mean(steps)) 
    %>% print)

g2 <- ggplot(data_plot2, aes(x = interval, y = mean_steps))
g2 + geom_line(color = "red") + facet_wrap(. ~ weekdays, nrow = 2) + labs(title = "Daily activity pattern by intervals", x = "Interval", y = "Average steps")
```
